---
title: "Sampling"
subtitle: "seeing what samples can do"
author: "Matthew Crump"
date: "2018/07/20 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["defaultb.css", "metropolisb.css", "metropolis-fontsb.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE, cache = TRUE)
```


# Big ideas for this course

1. Psychology interpets patterns in data to draw conclusions about psychological processes

--

2. Chance can produce "patterns" in data

--

3. **Problem**: How can we know if the pattern is real, or simply a random accident produced by chance

---

# Issues for this class

1. **Sampling from distributions**

2. **Normal distributions and central limit theorem**

3. **Sampling distributions and statistics**

---

class: pink, center, middle, clear

# Samples and populations

---

# Samples and populations

- Population: A defined set of things

- Sample: a subset of the population

---

# Random Sampling

- A process for generating a sample (taking things from a population)

--
- Random samples ensure that each value in a sample is drawn **independently** from other values

--
- all values in the population have a chance of being in the sample

---

# Example: Sampling heights of people

Let's say we wanted to know something about how tall people are. We can't measure the entire population (it's too big). So we take a sample.

--

What would happen if:

1. We only measured really tall people (biased sample)

--

2. We randomly measured a bunch of people?


---

# Population statistics

Populations have statistics. For example,

The population of all people has:

1. A distributions of heights
2. The distribution has a mean (mean height of all people)
3. The distribution has a standard deviation

---

# The population problem

In the real world, we usually do not have all of the data for the entire population.

So, we never actually know:

1. The population distribution
2. The population mean
3. The population standard deviation, etc.


---

# The sampling solution

Unknown: The population

Solution: Take a sample of the population

1. Samples will tend to look the population they came from, especially when sample-size (N) is large.

2. We can use the sample to **estimate** the population.

---

# The sampling problem

We take samples, and use them to estimate things. This works well when we have large, representative samples. 

But, how do we know if the sample we obtained is "normal", or happens to be "weird"?

Solution: We need to learn how the process of sampling works. We can use R to simulate the process of sampling. Then we can see how samples behave.

---

# Samples become populations

- As sample-size increases, the sample becomes more like the population. 

- As sample N approaches the population N, the sample becomes the population.

---

# Law of large numbers

- As sample-size increases, properties of the sample become more like properties of the population

Example:

- As sample-size increases, the mean of the sample becomes more like the mean of the population

---

# Simulation: Population mean=100

```{r}
save_mean<-c()
for(i in 1:1000){
  save_mean[i]<-mean(rnorm(i,100,25))
}
df<-data.frame(sample=1:1000,save_mean)

library(ggplot2)
ggplot(df, aes(x=sample, y=save_mean))+
  geom_point()+
  geom_line()+
  theme_classic(base_size = 12)+
  ggtitle("Sample mean approaches 100 \n
          as N increases")


```



---

# Sampling distributions

- 

---

# Next class: Sampling

1. Thursday, September 27th: We explore ideas about samples, populations, and distributions

---

# Reminder

1. Quiz 3 for this week is due today Tuesday, September, 25th end of day (11:59pm).

2. Quiz 4 for Distributions and Sampling will be posted by Wednesday, Due Next Tuesday (11:59 pm)


