---
title: "ANOVA"
author: "Matt"
date: "4/1/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,message=FALSE)
```

## one-way ANOVA

Create some fake data for three groups. n=5, all data sampled from a normal distribution with mean = 0, sd =1.
```{r}
A <- rnorm(n=5,mean=0,sd=1)
B <- rnorm(n=5,mean=0,sd=1)
C <- rnorm(n=5,mean=0,sd=1)
DV <- c(A,B,C)
groups <- rep(c("A","B","C"), each=5)
df <- data.frame(groups,DV)

knitr::kable(df)
```

### print ANOVA summary and graph

```{r}
library(ggplot2)
library(dplyr)

plot_df <- df %>%
            group_by(groups) %>%
            summarise(means=mean(DV))
ggplot(plot_df, aes(x=groups, y=means, fill=groups))+
  geom_bar(stat="identity")

summary(aov(DV~groups,df))
```

### repeat a few times

```{r}

for(i in 1:5){
  
A <- rnorm(n=5,mean=0,sd=1)
B <- rnorm(n=5,mean=0,sd=1)
C <- rnorm(n=5,mean=0,sd=1)
DV <- c(A,B,C)
groups <- rep(c("A","B","C"), each=5)
df <- data.frame(groups,DV)

plot_df <- df %>%
            group_by(groups) %>%
            summarise(means=mean(DV))

print(ggplot(plot_df, aes(x=groups, y=means, fill=groups))+
  geom_bar(stat="identity"))

print(summary(aov(DV~groups,df)))

}

```

# F-distribution

The null F-distribution, are the values of F you **could** have found, if all of your samples in each group are coming from the same normal distribution. We can compute this by simulation (pretending to run the experiment 1,000 times)

```{r}
save_F<-c()
for(i in 1:1000){
A <- rnorm(n=5,mean=0,sd=1)
B <- rnorm(n=5,mean=0,sd=1)
C <- rnorm(n=5,mean=0,sd=1)
DV <- c(A,B,C)
groups <- rep(c("A","B","C"), each=5)
df <- data.frame(groups,DV)

aov_sum <- summary(aov(DV~groups,df))
save_F[i]<-aov_sum[[1]]$`F value`[1]
}

```

## plotting the F-distribution

```{r}
plot_df <- data.frame(sims=1:1000,
                      save_F)

critical_f<-sort(save_F)[950]

ggplot(plot_df, aes(x=save_F))+
  geom_histogram()+
  geom_vline(xintercept=critical_f)
print(critical_f)
```



