---
title: "Data Sim 1"
author: "Matt"
date: "11/29/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simulating a t-test (paired sample)

Simulates a single experiment (assumes the null is true)

```{r}
n <- 10
sim_sd <- 50
mean_A <- 100
mean_B <- 100
A <- rnorm(n,mean_A,sim_sd)
B <- rnorm(n,mean_B,sim_sd)
t.test(A,B, paired=T)

# find cohen's d
differences <- A-B
mean_differences <- mean(differences)
sd_differences <- sd(differences)
cohens_d <- abs(mean_differences)/sd_differences
cohens_d

```


# Simulating a t-test (paired sample)

Simulating multiple t-tests of the same experiment (assumes the null is true)

```{r}
n <- 10
sim_sd <- 50
mean_A <- 100
mean_B <- 100

num_to_sim <- 1000
save_p <- c()
for(i in 1:num_to_sim){
  A <- rnorm(n,mean_A,sim_sd)
  B <- rnorm(n,mean_B,sim_sd)
  save_p[i] <- t.test(A,B, paired=T)$p.value
}

# plot histogram of p-values
hist(save_p)

# find proportion of ps less than alpha
length(save_p[save_p<.05])/num_to_sim

```


# Simulated Power analysis

Simulating multiple t-tests of the same experiment. 

Set `sim_sd` to 1. Set `mean_A` to 0. Set `mean_B` to the size of effect you want to simulate (d). 

Power is the proportion of simulations where p < alpha (0.05).

Note: Increase N to see the influence on power

```{r}
n <- 10
sim_sd <- 1
mean_A <- 0
mean_B <- 0.2

num_to_sim <- 1000
save_p <- c()
for(i in 1:num_to_sim){
  A <- rnorm(n,mean_A,sim_sd)
  B <- rnorm(n,mean_B,sim_sd)
  save_p[i] <- t.test(A,B, paired=T)$p.value
}

# plot histogram of p-values
hist(save_p)

# find proportion of ps less than alpha
power <- length(save_p[save_p<.05])/num_to_sim
power

```

# p-distribution when d is large

```{r}
n <- 10
sim_sd <- 1
mean_A <- 0
mean_B <- 1

num_to_sim <- 1000
save_p <- c()
for(i in 1:num_to_sim){
  A <- rnorm(n,mean_A,sim_sd)
  B <- rnorm(n,mean_B,sim_sd)
  save_p[i] <- t.test(A,B, paired=T)$p.value
}

# plot histogram of p-values
hist(save_p)

# find proportion of ps less than alpha
power <- length(save_p[save_p<.05])/num_to_sim
power

```


# Simulating power for different Ns

d=.01

```{r}
n <- c(10000,50000,100000)
sim_sd <- 1
mean_A <- 0
mean_B <- 0.01
num_to_sim <- 1000

power<-c()
for(j in 1:length(n)){
  save_p <- c()
  for(i in 1:num_to_sim){
    A <- rnorm(n[j],mean_A,sim_sd)
    B <- rnorm(n[j],mean_B,sim_sd)
    save_p[i] <- t.test(A,B, paired=T)$p.value
  }
  power[j] <- length(save_p[save_p<.05])/num_to_sim
}

# results in table
sim_results <- data.frame(n,power)
knitr::kable(sim_results)

# results in plot
library(ggplot2)
ggplot(sim_results, aes(x=n,y=power))+
  geom_point()+
  geom_line()+
  theme_classic(base_size=20)+
  ggtitle("Simulated Power analysis (d=.2)")

```



# Simulating power for different Ns

d=.5

```{r}
n <- c(10,20,30,40,50,60,70,80,90,100,200,500)
sim_sd <- 1
mean_A <- 0
mean_B <- 0.5
num_to_sim <- 1000

power<-c()
for(j in 1:length(n)){
  save_p <- c()
  for(i in 1:num_to_sim){
    A <- rnorm(n[j],mean_A,sim_sd)
    B <- rnorm(n[j],mean_B,sim_sd)
    save_p[i] <- t.test(A,B, paired=T)$p.value
  }
  power[j] <- length(save_p[save_p<.05])/num_to_sim
}

# results in table
sim_results <- data.frame(n,power)
knitr::kable(sim_results)

# results in plot
library(ggplot2)
ggplot(sim_results, aes(x=n,y=power))+
  geom_point()+
  geom_line()+
  theme_classic(base_size=20)+
  ggtitle("Simulated Power analysis (d=.2)")

```


# Simulating power for different Ns

d=.8

```{r}
n <- c(10,20,30,40,50,60,70,80,90,100,200,500)
sim_sd <- 1
mean_A <- 0
mean_B <- 0.5
num_to_sim <- 1000

power<-c()
for(j in 1:length(n)){
  save_p <- c()
  for(i in 1:num_to_sim){
    A <- rnorm(n[j],mean_A,sim_sd)
    B <- rnorm(n[j],mean_B,sim_sd)
    save_p[i] <- t.test(A,B, paired=T)$p.value
  }
  power[j] <- length(save_p[save_p<.05])/num_to_sim
}

# results in table
sim_results <- data.frame(n,power)
knitr::kable(sim_results)

# results in plot
library(ggplot2)
ggplot(sim_results, aes(x=n,y=power))+
  geom_point()+
  geom_line()+
  theme_classic(base_size=20)+
  ggtitle("Simulated Power analysis (d=.2)")

```
